Title here

### 2016-12-05 portfolio/text/ms.Rmd
Rework introduction and start cleaning up paper

A fundamental challenge for individuals who rely on the extraction of natural
resources for their livelihood is income variability.
Fishers, loggers, farmers, and miners, among others,
face income uncertainty driven by a mix of
environmental, biological, and economic variability.
A key component underlying the sustainability
of these industries is identifying both
how regulatory actions and individual behaviour affect income variability
and by extension the economic stability and
sustainable use of natural resources.
For commercial fisheries, recent work has identified that
targeting a diversity of species can reduce income variability.
However, any diversity-variability relationship
is embedded in a regulatory framework that restricts individual ability
to unilaterally change fishing behaviour.

Here we identify three primary axes by which fishers in Alaska,
one of the world's most productive fishery regions, can
potentially diversify fishing activities and
investigate how individual actions along each diversity axis affect
both revenue and revenue variability.
We construct a novel statistical model to simultaneously
assess the value to individual fishers of choosing to
(1) own permits that allow targeting a diverse fish species assemblage,
(2) diversify by owning multiple permits, or
(3) catch a greater species diversity within the constraints of a given permit.

We show contrasting effects of diversity on
revenue and revenue variability along the three axes.
First, we show that fishing permits with
higher species diversity reduces variability
(`r h1[[2]]`% [`r h1[[1]]`--`r h1[[3]]`% 95% CI] reduction per unit
increase in mean effective species diversity).
Next, we show that fishing an additional permit
is associated with higher revenue
and lower variability (e.g.,
`r 100-spoke[[2]]`% [`r 100-spoke[[3]]`--`r 100-spoke[[1]]`% IQR] lower
expected standard deviation
for halibut plus another permit vs. single halibut permits).
Finally, we show that increasing species diversity within
the constraints of a permit set has a fishery-dependent effect on revenue and
is usually (`r g2_dens`% probability)
associated with increased uncertainty in revenue the following year.

Thus, while there are clear benefits
to some axes of diversification, others are less effective.
Importantly, the most effective measure for
increasing revenue and decreasing risk ---
purchasing additional permits to participate in additional fisheries ---
was also the most costly.
At the same time, Alaskan commercial fisheries
have been increasingly focussed on
single-permit (often salmon) fishing strategies,
with single-permit fishers earning
more than multiple-permit fishers
for the first time in 2013.
Our results suggest that the benefit of species diversity
for revenue variability
is mostly an across-permit effect, not
a benefit that can be easily harnessed by an individual permit holder.
Effective methods of diversification ---
purchasing additional permits or switching permits ---
are costly and often limited by regulations such as limited entry
programs and individual transferable quotas (ITQs).
Take home ...

@@

Extra:

- there is often limited room for intentional diversification within a permit
- in fact, specializing can be profitable within some fisheries
- Eric: analog here in farming literature is that diversification
  for small scale farmers generally good, except when they
  specialize on high demand/high value crops (tobacco, cocoa).
  Seems like benefits of diversification are also smaller for
  large-scale industrial farmers.

\noindent
Main message:
Diversified portfolios of fishing permits and species reduce risk for fishermen,
but changing species diversity can be risky, costly, and limited by regulations.

@@

<!-- Introduction, Methods, Results, Discussion, Conclusion:
<!-- 600, 900, 500, 800, 200, -->

<!-- write out each paragraph’s main message in 15 words or less -->
<!-- circulate skeleton ... this is the time to fix problems with flow, -->
<!-- ideas/content, main message -->

@@

Managing natural resources and sustaining the livelihoods of those that depend on their extraction is difficult. These professions tend to have high year-to-year variability in income [@mishra2002; @sethi2012b]. In addition to economic risk, such as changes in prices or demand, individuals dependent on natural resources are also subject to biological and environmental variability. For example, drought and flooding are a major source of risk to agricultural food security and farmers' incomes [@gregory2005], and catastrophic disease outbreaks or wildfires increase risk for the logging industry [@kirilenko2007].

Individuals who rely on natural resources for income develop strategies to reduce this variability in income. For example, farmers may rotate their crops to buffer short-term climate variation and long-term environmental trends [@zentner1988]. Similarly, loggers maintain a diverse mix of tree species, age structure, and harvest in geographical areas to buffer against disease outbreak, wildfires, and price variation REF.

In many of these socio-ecological systems, otherwise well-intentioned regulations can limit the actions available to individuals to reduce income variability, or incentivize against diversification. For example, in terrestrial agriculture, crop subsidies in the US may incentivize some farms to specialize [e.g. dairy, grains, soybeans;  @mishra2002]. In other instances, external market forces (prices, demand) may discourage the adoption of diverse strategies and help promote specialization [@bradshaw2004]. In fisheries, a series of sequential changes around the world have generally limited participation and led to increased specialization [@hilborn2003a].

Commercial fishers present an interesting example of diversification, given the extreme risk that some individuals experience, particularly relative to terrestrial agriculture (@sethi2012b). First, the majority of insurance programs available to terrestrial farmers are not available to commercial fishermen [e.g. crop insurance programs; @kasperski2013]. Second, as both fish and fishers are spatially mobile, equating fishers to farmers would be possible if the farmers’ field became an amorphous blob that changed shape and moved unpredictably day to day. Fishermen still have the ability to lower risk, however. While the days of open, unregulated fisheries in most developed countries have passed (fishers must own a license to participate in virtually all fisheries), fishers can fish multiple species as permitted by a license, own multiple licenses, or diversify spatially and fish in multiple regions.

Despite these ways fishers can diversify, the effect of adopting various diversification strategies for improving revenue and reducing revenue variability remains uncertain. Furthermore, theory and empirical studies both suggest there can be substantial benefits to specialization (REFS). The majority of papers that examine effects of diversification on revenue do so at aggregate levels (e.g. households, counties, communities; REFS), rather than focus on individuals. Because of this, key uncertainties remain about the fundamental relationship between diversity, fishing revenue, and fishing revenue variability at the level of the individual.

One of the most well-studied examples of diversification are the state and federal fisheries in Alaska. These provide a unique opportunity to understand the role of individuals' strategies in determining income and income variability for those whose livelihood depends on the extraction of natural resources. Fisheries in Alaska are among the highest-valued fisheries in the world with 6 billion landed pounds [TODO kg] valued at nearly 2 billion USD annually [@nmfs2016]. The industry supports a range of fishing fleets ranging from small, open, single-operator vessels to multi-million-dollar state-of-the-art ships with large crews. Like much of the rest of the world [@hilborn2003a], the days of open, unregulated fisheries in most developed countries have passed --- fishers must own a license to participate in virtually all fisheries. And while such regulations have proven helpful in reducing or eliminating overfishing, they do limit the available strategies to reduce income variability.

Using a unique dataset collected over 30 years (1985--2014) collected on individuals, rather than fleets or vessels, we examine how revenue variability of individuals changes through time, and is affected by diversification. The benefit of using individuals rather than vessels is that we are able to track fishers that switch fishery sectors over time, or alter strategies and use multiple vessels. We develop and apply a novel statistical approach for understanding the effect of diversity on both revenue and risk that builds on results from the vessel [@kasperski2013], fleet [@sethi2012b], and community level variability [@sethi2014]. To our knowledge, this is the first application of this type of model in any natural resource field that jointly explores the effect of diversity on revenue and risk at the individual level and the first considers the multiple axes of diversity simultaneously.

First, we summarize trends in diversification across fishing fleets in Alaska. Next, we identify three main axes by which individuals may diversify the species they target [Eric: I think these kind of bleed together, perhaps make more clear or reduce to 2]:
(1) owning permits that allow targeting a diversity of fish species (e.g. miscellaneous groundfish or salmon permits),
(2) purchasing additional permits (the majority of permits in Alaska are single species – halibut, herring, etc),
or (3) catching a greater species diversity within the constraints of a given permit.
*perhaps we also need to have a sentence or two on the implicit constraints of
geography and gear / vessel size on species diversification*

First, the effect of changing species diversity
for individual permit holders on a year-by-year basis is unknown ---
existing approaches have worked at the vessel [@kasperski2013],
fleet [@sethi2012b], or community level [@sethi2014],
and condensed time series into single metrics of variability or risk.
Second, the effect of changing species diversity
on revenue itself is unknown ---
existing approaches have focused solely on variability or risk.
Third, it is unknown if the approach to
diversifying species fished affects the relationship between
species diversity and revenue variability ---
a fishery permit holder can purchase
an additional permit to target other species,
pick a permit that inherently targets many species,
or increase species diversity
within the constraints of his or her existing permits.
Many permit holders may be limited to the last approach
given that the first two may be costly and limited by regulations,
but it is unclear if the benefits and costs are the same.
Here, we address these questions by
simultaneously estimating the effects of
these three approaches to species diversification
for Alaskan fishery permit holders on
annual revenue and revenue variability.
Our model makes use of entire time series,
controls for effort and latent time effects,
and allows the effect of species diversity
to vary by fishing strategy (sets of similar permits)
in a hierarchical model.

@@

### Fig. \ref{fig:ts} - limited entry + movement towards salmon and specialization in Alaska

@@

### Fig. \ref{fig:across} - across strategies

@@

### Fig. \ref{fig:spokes} - add permits

@@

### Fig. \ref{fig:within} - within strategies

@@

# Notes

## @kasperski2013
- vessel-level CFEC data
- dome-shaped result
- Herfindahl-Hirschman Index/HHI (Simpson)
- diversity decreasing for those still fishing in 2010, but not otherwise

## @sethi2012b
- risk measures to 90 Alaskan fishing fishery/fleets
- revenue more variable than catch (+ price = extra variability)
- large catches, big fleets = lower risk

## @sethi2010b
- risk management for fisheries review

## @sethi2012
- equal-share catch-pooling cooperatives in king and snow crab
- "variability reduction depends on the degree of correlation amongst
  participants’ catches"

## @sethi2014
- community-level CFEC data
- number of fisheries strongest predictor of (lower) CV/CVaR
- also simpson index
- suggest barriers to commercial entry increase community risk by restricting
  access to 'variety of fisheries'
- some communities have limited ability to diversify; may need to diversify
  outside of fishing

[@perruso2005]
[@hilborn2001]
[@edwards2004]
[@trenkel2013]
[@mumford2009]
[@oostenbrugge2002]
[@baldursson1997]
[@jardine2015]
[@sanchirico2008]

>>>>

A fundamental challenge for individuals who rely on the extraction of natural
resources for their livelihood is income variability.
Fishers, loggers, farmers, and miners, among others,
face income uncertainty driven by a mix of
environmental, biological, and economic variability.
A key component underlying the sustainability
of these industries is identifying both
how regulatory actions and individual behaviour affect income variability
and by extension the economic stability and
sustainable use of natural resources.
For commercial fisheries, recent work has identified that
targeting a diversity of species can reduce income variability.
However, any diversity-variability relationship
is embedded in a regulatory framework that restricts individual ability to
change fishing behaviour. Here we identify three primary axes by which fishers
in Alaska, one of the world's most productive fishery regions, can potentially
diversify fishing activities and investigate how individual actions along each
diversity axis affect both revenue and revenue variability. We construct a novel
statistical model to simultaneously assess the value to individual fishers of
choosing to (1) own permits that allow targeting a diverse fish species
assemblage, (2) diversify by owning multiple permits, or (3) catch a greater
species diversity within the constraints of a given permit.
We show contrasting effects of diversity on
revenue and revenue variability along the three axes.
First, we show that fishing permits with
higher species diversity reduces variability
(`r h1[[2]]`% [`r h1[[1]]`--`r h1[[3]]`% 95% CI] reduction per unit
increase in mean effective species diversity).
Next, we show that fishing an additional permit
is associated with higher revenue
and lower variability (e.g.,
`r 100-spoke[[2]]`% [`r 100-spoke[[3]]`--`r 100-spoke[[1]]`% IQR] lower
expected standard deviation
for halibut plus another permit vs. single halibut permits).
Finally, we show that increasing species diversity within
the constraints of a permit set has a fishery-dependent effect on revenue and
is usually (`r g2_dens`% probability)
associated with increased uncertainty in revenue the following year.
Thus, while there are clear benefits
to some axes of diversification, others are less effective.
Importantly, the most effective measure for
increasing revenue and decreasing risk ---
purchasing additional permits to participate in additional fisheries ---
was also the most costly.
At the same time, Alaskan commercial fisheries
have been increasingly focussed on
single-permit (often salmon) fishing strategies,
with single-permit fishers earning
more than multiple-permit fishers
for the first time in 2013.
Our results suggest that the benefit of species diversity
for revenue variability
is mostly an across-permit effect, not
a benefit that can be easily harnessed by an individual permit holder.
Effective methods of diversification ---
purchasing additional permits or switching permits ---
are costly and often limited by regulations such as limited entry
programs and individual transferable quotas (ITQs).
Take home ...

@@

<!-- Extra: -->

<!-- - there is often limited room for intentional diversification within a permit -->
<!-- - in fact, specializing can be profitable within some fisheries -->
<!-- - Eric: analog here in farming literature is that diversification -->
<!--   for small scale farmers generally good, except when they -->
<!--   specialize on high demand/high value crops (tobacco, cocoa). -->
<!--   Seems like benefits of diversification are also smaller for -->
<!--   large-scale industrial farmers. -->

<!-- \noindent -->
<!-- Main message: -->
<!-- Diversified portfolios of fishing permits and species reduce risk for fishermen, -->
<!-- but changing species diversity can be risky, costly, and limited by regulations. -->

@@

<!-- Goal: Introduction, Methods, Results, Discussion, Conclusion:
<!-- 600, 900, 500, 800, 200, -->

@@

<!-- Papers to cite [@smith1986] -->

Managing natural resources and sustaining the livelihoods of those that depend
on their extraction is difficult. These professions tend to have high
year-to-year variability in income [@mishra2002; @sethi2012b]. In addition to
economic risk, such as changes in demand, individuals dependent on natural
resources are also subject to biological and environmental variability. For
example, drought and flooding are a major source of risk to agricultural food
security and farmers' incomes [@gregory2005], and catastrophic disease outbreaks
and wildfires increase risk for the logging industry [@kirilenko2007].

Individuals who rely on natural resources for income develop strategies to
reduce income variability. For example, farmers may rotate their crops to buffer
short-term climate variation and long-term environmental trends [@zentner1988].
Similarly, loggers maintain a diverse mix of tree species, age structure, and
harvest in geographical areas to buffer against disease outbreak, wildfires, and
price variation REF. However, otherwise well-intentioned regulations can limit
the actions available to individuals to reduce income variability, or
incentivize against diversification. For example, in terrestrial agriculture,
crop subsidies in the US may incentivize some farms to specialize [e.g. dairy,
grains, soybeans; @mishra2002]. In other instances, external market forces
(prices, demand) may discourage the adoption of diverse strategies and help
promote specialization [@bradshaw2004].

Commercial fishers present an interesting example of diversification, given the
extreme risk that some individuals experience, particularly relative to
terrestrial agriculture (@sethi2012b). The majority of insurance programs
available to terrestrial farmers (e.g. crop insurance programs) are not
available to commercial fishers [@kasperski2013] and both fish and fishers are
spatially mobile. However, fishers still have the ability to lower risk. They
can, for example, target multiple species as permitted by a license, own
multiple licenses, or diversify spatially and fish in multiple regions. The
effectiveness of these diversification strategies for improving revenue and
reducing revenue variability remains uncertain. Furthermore, theory and
empirical studies both suggest there can be substantial benefits to
specialization (REFS). The majority of papers that examine effects of
diversification on revenue do so at aggregate levels (e.g. households, counties,
communities; REFS), rather than focus on individuals. Because of this, key
uncertainties remain about the fundamental relationship between diversity,
fishing revenue, and fishing revenue variability at the level of the individual.

The state and federal fisheries of Alaska, provide a unique opportunity to
understand the role of individuals' strategies in determining income and income
variability for those whose livelihood depends on the extraction of natural
resources. Fisheries in Alaska are among the highest-valued fisheries in the
world with 2.7 billion kg landed biomass valued at nearly 2 billion USD annually
[@nmfs2016]. The industry supports a range of fishing fleets ranging from small,
open, single-operator vessels to multi-million-dollar state-of-the-art ships
with large crews. Like much of the rest of the world [@hilborn2003a], the days
of open, unregulated fisheries in most developed countries have passed ---
fishers must own a license to participate in virtually all fisheries. And while
such regulations have proven helpful in reducing or eliminating overfishing,
they limit the available strategies to reduce income variability. For example,
most permits limit fishing to a single species, diversifying by purchasing
additional fishing permits can be expensive (e.g. ~\$XX for king crab), and some
fisheries limit the entry of new participants through individual fishing quotas
(IFQs) (REF).

Using a unique dataset collected over 30 years (1985--2014) on individuals,
rather than fleets or vessels, we examine how revenue and revenue variability of
individuals is affected by diversification. By tracking individuals rather than
vessels we can track people that switch fishery sectors over time, or alter
strategies and use multiple vessels. We develop and apply a novel statistical
approach for understanding the effect of diversity on both revenue and revenue
variability that builds on results from vessel [@kasperski2013], fleet
[@sethi2012b], and community-level variability [@sethi2014]. Our model controls
for fishing effort and latent time effects (e.g. changing environmental
conditions), simultaneously estimates the effect of species diversity on
individual revenue and revenue variability, and estimates effects both within
and across fishing strategies (sets of similar licenses or permits). To our
knowledge, this is the first application of this type of model in any natural
resource field.

@@

**Fig. \ref{fig:ts} - limited entry + movement towards salmon and specialization
in Alaska**

@@

**Fig. \ref{fig:across} - across strategies**

@@

**Fig. \ref{fig:spokes} - add permits**

@@

**Fig. \ref{fig:within} - within strategies**

@@

<!-- # Notes -->

<!-- ## @kasperski2013 -->
<!-- - vessel-level CFEC data -->
<!-- - dome-shaped result -->
<!-- - Herfindahl-Hirschman Index/HHI (Simpson) -->
<!-- - diversity decreasing for those still fishing in 2010, but not otherwise -->

<!-- ## @sethi2012b -->
<!-- - risk measures to 90 Alaskan fishing fishery/fleets -->
<!-- - revenue more variable than catch (+ price = extra variability) -->
<!-- - large catches, big fleets = lower risk -->

<!-- ## @sethi2010b -->
<!-- - risk management for fisheries review -->

<!-- ## @sethi2012 -->
<!-- - equal-share catch-pooling cooperatives in king and snow crab -->
<!-- - "variability reduction depends on the degree of correlation amongst -->
<!--   participants’ catches" -->

<!-- ## @sethi2014 -->
<!-- - community-level CFEC data -->
<!-- - number of fisheries strongest predictor of (lower) CV/CVaR -->
<!-- - also simpson index -->
<!-- - suggest barriers to commercial entry increase community risk by restricting -->
<!--   access to 'variety of fisheries' -->
<!-- - some communities have limited ability to diversify; may need to diversify -->
<!--   outside of fishing -->

<!-- [@perruso2005] -->
<!-- [@hilborn2001] -->
<!-- [@edwards2004] -->
<!-- [@trenkel2013] -->
<!-- [@mumford2009] -->
<!-- [@oostenbrugge2002] -->
<!-- [@baldursson1997] -->
<!-- [@jardine2015] -->
<!-- [@sanchirico2008] -->
